---
title: "Introduction to 2017 Statistics Methods Forum Data Challenge"
author: "Eric Polley"
date: "Oct. 18th, 2017"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

The focus this year is the estimation of a causal treatment effect from a retrospective study\footnote{Partially motivated by the Atlantic Causal Inference Data Challenge \url{http://causal.unc.edu/acic2017}}

A dataset with 400 patients will be provided with a binary treatment, a continuous outcome of interest, and a set of potential confounders or covariates. The primary goal is to estimate the average treatment effect and provide a 95\% confidence interval for the estimate.

Details for the Challenge available on Github: \url{https://github.com/ecpolley/Data_Challenge_2017}

## Outline

Will continue for the next two regular Statistical Methods Forums 
(2-3pm central)

* Oct. 18th: Introduction to the data challenge and dataset
* Nov. 15th: Group discussion and Q&A session
* Dec. 18th, 5:00pm local: Team submissions deadline (If team is across sites, depends who sends the results)
* Dec. 20th: Final results and team scores, and discussion of methods used

## Team Science

* Participants are encouraged to work in teams\newline (N $\in$ (1, 2, \ldots, 10))
* Opportunity to learn from each other and work with people outside usual team
* Data is publicly available, so is available outside Mayo 
* If you would like help forming a team, email Eric Polley, Kristin Mara, or Sara Fett
* Teams are responsible for creating a team name, and may submit up to 3 estimates, with the last submission being the official one
* If you are participating, please let us know in case we have any Data Challenge announcements

## Overview of Dataset

* CSV file available on Github, \url{https://github.com/ecpolley/Data_Challenge_2017}
* 400 independent observations (rows)
* Y: Continuous outcome of interest
* A: Binary treatment indicator
* W1-W25: Baseline measured variables, includes all confounders
* ID: Individual id number
* No missing data, No hidden messages in 10-dimension space\footnote{See \url{https://github.com/ecpolley/CSMF_Data_Challenge/blob/master/Biomarkers.R}}

## Overview of Dataset

```{r loadData, echo = TRUE}
# link to data on GitHub page if not available
if(file.exists("Data.csv")) {
  Dat <- read.csv("Data.csv")
} else {
  urlfile <- "https://raw.githubusercontent.com/ecpolley/
    Data_Challenge_2017/master/Data.csv"
  download.file(urlfile, destfile = "Data.csv")
  Dat <- read.csv("Data.csv")
}
dim(Dat)
```

## Primary Objective

The primary goal is to estimate the average treatment effect (ATE). We can define the values $Y(0)$ and $Y(1)$ to be the possibly counterfactual outcome values had the patient been given treatment 0 and treatment 1, respectively. In the dataset, the observed value $Y$ is:
\[
Y_i = (1 - A_i)Y_i(0) + A_iY_i(1)
\]
The parameter of interest is the ATE:
\[
\psi = E(Y(1) - Y(0))
\]
and provide a 95% confidence interval for the estimate.

Teams scores based on distance between estimate and true value, and the width of teh confidence interval. A penalty will be added if the true value is outside the interval.

## Primary Objective

Team results can be emailed to Eric (\url{Polley.Eric@Mayo.edu}), with the following:

1) Team members
2) Team name
3) ATE estimate
4) Lower and Upper confidence limits


## Secondary Objective

The secondary goal is to estimate the individual treatment effect for all 400 samples:
\[
\psi_i = Y_i(1) - Y_i(0),  i \in 1, \ldots, N
\] 

The mean squared error with the true individual treatment effect will be computed (*i.e.* precision in estimation of heterogeneous effects), along with the concordance of the sign ($+/-$) of the effect.

## Secondary Objective

Team results for the optional secondary objective can be emailed to Eric (\url{Polley.Eric@Mayo.edu}) with the following:

1) Team members
2) Team name
3) Text file with 2 columns: ID variable and predicted individual treatment effect

## Example (Ignore Confounding)

```{r ttest, echo = TRUE}
t.test(Y~A, data = Dat)
with(Dat, mean(Y[A == 1]) - mean(Y[A == 0]))
```

## Extensions

* As a team should discuss different methods to adjust for confounding (*e.g.* Regression model, Machine Learning, IPTW, propensity score matching, Targeted MLE, etc.)
* Depending on selection of method for estimation, need to estimate confidence interval (*e.g.* closed form approximation, resampling, etc.)

## Questions?
